market-justice-preferences

### 2024-04-23 paper.qmd
update references
faltan aun

---
title: "market-justice-preferences"
author: "Equipo EDUMER"
format: html
editor: visual
bibliography: "input/bib/just-ineq-merit.bib"
---

@@

Given that the school environment has an important focus on performance, achievement and acknowledgment, meritocracy has been one of the principal concepts used for understanding and even for justifying performance differences among students. Meritocracy is a distributive system based on the belief that people should be rewarded and promoted based on their abilities, knowledge, and achievements (Young, 1958). It is often seen as a way to create equal opportunities and fairness, as individuals can rise to positions of power and influence based on their own merit rather than their background or connections. However, some argue that meritocracy can actually lead to tolerating or even justifying social inequalities, as it can create a hierarchy where those who already have resources and advantages are more likely to succeed. In this regard, a great deal of academic research about meritocracy delves into the assessment of to what extent rewards and privileges in society are related to merit, emphasizing the so-called unfulfillable promise of meritocracy (Mijs, 2016). Complementing this agenda, a second and emerging research area deals with subjective aspects of meritocracy, such as perceptions and beliefs.

@@

The perception of meritocracy refers to how individuals view and understand the concept of meritocracy in their own society (Duru-Bellat & Tenret, 2012; Castillo et al., 2019). This perception can vary greatly depending on individual experiences, social, economic, and cultural background. Some people may see meritocracy as a fair and just system that allows anyone to succeed based on their abilities and hard work. In contrast, others may view it as a myth or a cover for existing power dynamics and inequality, serving to maintain and even reinforce inequality (Lampert, Mijs, etc). Based in this last perspective, we argue that individuals with a higher perception of meritocracy will show a larger justification of social inequalities, as individual achievement would be seen as rewarded and social policies as less necessary (Batruch et al, 2021).

Most of the research that has related meritocratic beliefs to inequality justification so far has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the \_hidden curriculum\_ (Chafel, 1997, https://doi.org/10.1002/J.2379-3988.1997.TB00004.X), whereby students learn about distributive norms in society and mechanisms of justification of social differences. Based on recent studies that relate school meritocracy to the justification of economic inequalities in the adult population \[\@batruch_belief_2022;\@wiederkehr_belief_2015\], the central hypothesis guiding this research is that school-age students with a higher perception of meritocracy - both at school and at the societal level - will show a larger justification of social inequalities, as individual achievement would be seen as appropriately rewarded and social mechanisms for correcting inequalities as less necessary (Batruch et al, 2021).

@@

Research on social stratification beliefs, which explore individual perceptions of who deserves what and why (Kluegel & Smith, 1987), highlights that people’s explanations and justifications of social inequality are closely tied to their judgments of deservingness. The influence of ideologies (Wegener & Liebig, 1995) and cultural schemas (Homan et al., 2017) is pivotal in shaping these explanations by offering symbolic representations that frame societal structures and expectations. While significant attention has been paid to wage inequality, income distribution, and payment differentials in the literature (Castillo, 2011; Evans et al., 2010; Jasso, 1999; Shariff et al., 2016),  there has been less examination of public beliefs about which life domains should be governed by market relations and even less about children’s acceptance or rejection of these market principles. This oversight is notable given the extensive encroachment of market logic into public goods, welfare policy, and social services over the past five decades (Centeno & Cohen, 2012; Harvey, 2007), affecting areas such as pensions, health services, and education. The justification of social inequality based on marketype criteria has been conceptualized as the individuals’ adherence to one specific justice evaluation, the market justice, that is, affording legitimacy to the allocation of goods and services based on prices and individuals’ ability to pay (Boltanski & Chiapello- 2007; Lane 1986; Streeck 2012).

@@

The question arises whether adults and children justify unequal access to welfare services based on market justice principles. Influenced by theories of policy feedback, which suggest that social welfare policies can reinforce (positive feedback) or undermine (negative feedback) previous policy trajectories (Fernandez and Jaime-Castillo 2013; Pierson 2000; Weaver 2010), citizens’ beliefs about market justice are likely also shaped by the institutional and social contexts they encounter. Indeed, the justification of inequality in access to essential services like education and health, based on one’s ability to pay, shows significant variation across countries, as demonstrated by international surveys, although they do not usually include children in their samples. For instance, the 2019 International Social Survey Program (ISSP) provides insights into how adults perceive inequality in accessing welfare services. Figure 1 illustrates the variation in agreement levels regarding whether it is just for individuals with higher incomes to purchase better education or healthcare. Notably, in 18 of the 29 countries analyzed, there is a greater justification for market inequality in healthcare access than in education. Nevertheless, the general sentiment typically ranges from 'somewhat unjust' to 'neither just nor unjust, with mixed feelings.

@@

While contemporary policy developments have increasingly embraced the marketization of social welfare in areas such as education, pensions, and health services, questions about public acceptance of these changes persist. Lindh’s (2015) analysis of ISSP 2009 data from 17 OECD countries reveals a general lack of support for market-based distribution of social services, suggesting widespread disapproval of market stratification of essential services. This finding is corroborated by Soler-Martínez and colleagues’ (2023) research from Latinobarómetro 2020 across 18 Latin American countries, where concerns about health and education access predominated over income inequality. These results indicate that reforms toward welfare marketization are typically driven by elite political decisions rather than grassroots demand.

Despite high-income inequality and limited social mobility in Latin America, there is a prevalent belief that individuals are solely responsible for their economic outcomes, a view that varies across the region (Bucca, 2016; Chong et al., 2008; Torche, 2014; Salgado and Castillo, 2023). The reliance on private welfare providers and widespread user fees (Molyneux, 2008) adds complexity to this context. Yet, research on children’s justification of market-based inequalities in accessing welfare services remains limited, especially in Latin America, highlighting a significant gap in understanding how younger generations view market-based access to welfare and whether these views are associated with their meritocratic beliefs.

@@

The concept of meritocracy frequently appears nowadays when analyzing cultural determinants of social inequalities. In general, it is mentioned as a value associated with justice, as it would link efforts and talents with rewards in an equitable manner. This normative sense is quite far from its original formulation by Michael Young (1958) in the satirical novel "The Rise of Meritocracy", where it ironically represented a mechanism for reproducing the inequalities of origin. The meritocratic ideal had remained relatively unchallenged until a series of recent publications turned into its potential consequences for maintaining social inequality. Perhaps one of the most recent sources in this line is Michael Sandel's "The Tyranny of Merit", where he strongly questions the implications of carrying out the principle of merit in societies that do not guarantee equal opportunities and that generate a feeling of scarce recognition and appreciation of those who receive lesser rewards: "In society's eyes, and perhaps also their own, their work no longer signified a valued contribution to the common good." (Sandel, pp).

Empirical research on meritocracy has increased along with the philosophical-normative discussion on meritocracy in recent years. Particularly from a sociological perspective, meritocracy has been used in research on social mobility to characterize societies with low mobility that threaten the meritocratic ideal (Goldthorpe). More recently, sociology and social psychology research has attended to the subjective aspects vis-a-vis beliefs in meritocracy. The label of beliefs in this realm covers a series of areas, such as attitudes, perceptions, and preferences (Castillo et al), whereby most of the link this subjective dimensions to individual socio-structural factors and  context-level determinants. For instance, some studies have analyzed how those with greater privileges believe more in meritocracy (Reynolds & Chan 2014), how greater economic inequality increases meritocratic beliefs (Mijs 2021), and how larger inequality affects meritocratic beliefs (Morris et al 2022). Based on these findings, a research agenda has been reinforced on the legitimizing role of meritocracy, in line with previous studies using the concept of a just world (Lerner, Dalbert) and the theory of system justification (Jost & Major).

How do meritocratic beliefs legitimize inequalities? Empirical studies have used experiments and surveys to address this question. For instance, the evidence suggests that just world beliefs correlate negatively with support for redistributive compensation systems (Frank et al., 2015). Conversely, individuals tend to support redistribution when they believe that the disadvantaged lack the opportunities to succeed (Evans & Kelley, 2018). Almås and colleagues (2020) found that in a relatively unequal society (the United States), the highly educated accept inequality significantly more than the less educated because they perceive inequality as justifiable owing to differences in productivity (i.e., merit), whereas in a relatively equal society (Norway), the less educated accept inequality more, but not significantly more than the highly educated because meritocratic values are less prevalent. Barr and Miller (2020) also addressed this triple interaction between the level of inequality in a society, the individual level of education, and the perceived origin of the disparity (either by luck or effort) to determine the extent to which inequality is accepted. They found that the interaction's mechanism varies depending on the compared societies. Finally, García-Sánchez and colleagues (2020), using data for 41 countries from the International Social Survey Programme (ISSP), found that the perceived size of the income gap correlated positively with support for progressive taxation. Still, this association was weaker among those who endorsed meritocratic and equal opportunity beliefs. In the same line, experimental research by Durante & Putterman (2009) subjects support less redistribution when the initial distribution is determined according to task performance.

Research about meritocratic beliefs at school age is rather scarce, leaving a wide research gap as schools are one of the primary socialization institutions where achievement based on merit explains success (Erivwo et al 2021). Nevertheless, it is possible to find some initial works focused on the area of distributive justice at school that are closely related to meritocratic beliefs, as the ones by Resh and Sabbagh (quotes). Using justice in grade obtention as a measure of distributive justice and meritocracy (Sabbagh et al 2006), they find for instance that a larger sense of distributive justice about grades is associated to higher socio-economic status (Resh 2010), have a positive effect on liberal democratic orientation and on trust in people and in formal institutions (Resh & Sabbagh 2013), and tend to refrain from violence and to engage to a greater extent in extra-curricular school activity and community volunteering (Resh & Sabbagh 2017).

>>>>

---
title: "market-justice-preferences"
author: "Equipo EDUMER"
format: pdf
editor: visual
bibliography: "input/bib/just-ineq-merit.bib"
link-citations: true
---

@@

Given that the school environment has an important focus on performance, achievement and acknowledgment, meritocracy has been one of the principal concepts used for understanding and even for justifying performance differences among students. Meritocracy is a distributive system based on the belief that people should be rewarded and promoted based on their abilities, knowledge, and achievements [@young_rise_1958]. It is often seen as a way to create equal opportunities and fairness, as individuals can rise to positions of power and influence based on their own merit rather than their background or connections. However, some argue that meritocracy can actually lead to tolerating or even justifying social inequalities, as it can create a hierarchy where those who already have resources and advantages are more likely to succeed. In this regard, a great deal of academic research about meritocracy delves into the assessment of to what extent rewards and privileges in society are related to merit, emphasizing the so-called unfulfillable promise of meritocracy [@mijs_stratified_2016]. Complementing this agenda, a second and emerging research area deals with subjective aspects of meritocracy, such as perceptions and beliefs.

@@

The perception of meritocracy refers to how individuals view and understand the concept of meritocracy in their own society [@duru-bellat_who_2012; @castillo_meritocracia_2019]. This perception can vary greatly depending on individual experiences, social, economic, and cultural background. Some people may see meritocracy as a fair and just system that allows anyone to succeed based on their abilities and hard work. In contrast, others may view it as a myth or a cover for existing power dynamics and inequality, serving to maintain and even reinforce inequality [@lampert_meritocratic_2013; @mijs_paradox_2021a]. Based in this last perspective, we argue that individuals with a higher perception of meritocracy will show a larger justification of social inequalities, as individual achievement would be seen as rewarded and social policies as less necessary [@batruch_belief_2022].

Most of the research that has related meritocratic beliefs to inequality justification so far has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the *hidden curriculum* [@chafel_schooling_1997], whereby students learn about distributive norms in society and mechanisms of justification of social differences. Based on recent studies that relate school meritocracy to the justification of economic inequalities in the adult population [@batruch_belief_2022; @wiederkehr_belief_2015], the central hypothesis guiding this research is that school-age students with a higher perception of meritocracy - both at school and at the societal level - will show a larger justification of social inequalities, as individual achievement would be seen as appropriately rewarded and social mechanisms for correcting inequalities as less necessary [@batruch_belief_2022].

@@

Research on social stratification beliefs, which explore individual perceptions of who deserves what and why [@kluegelBeliefsInequalityAmericans1987], highlights that people’s explanations and justifications of social inequality are closely tied to their judgments of deservingness. The influence of ideologies [@wegenerDominantIdeologiesVariation1995] and cultural schemas [@homanBeingBecomingPoor2017] is pivotal in shaping these explanations by offering symbolic representations that frame societal structures and expectations. While significant attention has been paid to wage inequality, income distribution, and payment differentials in the literature [@castilloLegitimacyInequalityHighly2011; ] (Castillo, 2011; Evans et al., 2010; Jasso, 1999; Shariff et al., 2016),  there has been less examination of public beliefs about which life domains should be governed by market relations and even less about children’s acceptance or rejection of these market principles. This oversight is notable given the extensive encroachment of market logic into public goods, welfare policy, and social services over the past five decades (Centeno & Cohen, 2012; Harvey, 2007), affecting areas such as pensions, health services, and education. The justification of social inequality based on marketype criteria has been conceptualized as the individuals’ adherence to one specific justice evaluation, the market justice, that is, affording legitimacy to the allocation of goods and services based on prices and individuals’ ability to pay (Boltanski & Chiapello- 2007; Lane 1986; Streeck 2012).

@@

The question arises whether adults and children justify unequal access to welfare services based on market justice principles. Influenced by theories of policy feedback, which suggest that social welfare policies can reinforce (positive feedback) or undermine (negative feedback) previous policy trajectories [@fernandez_positive_2013] (Fernandez and Jaime-Castillo 2013; Pierson 2000; Weaver 2010), citizens’ beliefs about market justice are likely also shaped by the institutional and social contexts they encounter. Indeed, the justification of inequality in access to essential services like education and health, based on one’s ability to pay, shows significant variation across countries, as demonstrated by international surveys, although they do not usually include children in their samples. For instance, the 2019 International Social Survey Program (ISSP) provides insights into how adults perceive inequality in accessing welfare services. Figure 1 illustrates the variation in agreement levels regarding whether it is just for individuals with higher incomes to purchase better education or healthcare. Notably, in 18 of the 29 countries analyzed, there is a greater justification for market inequality in healthcare access than in education. Nevertheless, the general sentiment typically ranges from 'somewhat unjust' to 'neither just nor unjust, with mixed feelings.

@@

While contemporary policy developments have increasingly embraced the marketization of social welfare in areas such as education, pensions, and health services, questions about public acceptance of these changes persist. @lindh_public_2015 analysis of ISSP 2009 data from 17 OECD countries reveals a general lack of support for market-based distribution of social services, suggesting widespread disapproval of market stratification of essential services. This finding is corroborated by Soler-Martínez and colleagues’ (2023) research from Latinobarómetro 2020 across 18 Latin American countries, where concerns about health and education access predominated over income inequality. These results indicate that reforms toward welfare marketization are typically driven by elite political decisions rather than grassroots demand.

Despite high-income inequality and limited social mobility in Latin America, there is a prevalent belief that individuals are solely responsible for their economic outcomes, a view that varies across the region [@buccaMeritBlameUnequal2016; @chongMysteryDiscriminationLatin2008; @torcheIntergenerationalMobilityInequality2014; @salgadoInequalityStratificationLatin2023]. The reliance on private welfare providers and widespread user fees (Molyneux, 2008) adds complexity to this context. Yet, research on children’s justification of market-based inequalities in accessing welfare services remains limited, especially in Latin America, highlighting a significant gap in understanding how younger generations view market-based access to welfare and whether these views are associated with their meritocratic beliefs.

@@

The concept of meritocracy frequently appears nowadays when analyzing cultural determinants of social inequalities. In general, it is mentioned as a value associated with justice, as it would link efforts and talents with rewards in an equitable manner. This normative sense is quite far from its original formulation by @young_rise_1958 in the satirical novel "The Rise of Meritocracy", where it ironically represented a mechanism for reproducing the inequalities of origin. The meritocratic ideal had remained relatively unchallenged until a series of recent publications turned into its potential consequences for maintaining social inequality. Perhaps one of the most recent sources in this line is Michael Sandel's "The Tyranny of Merit", where he strongly questions the implications of carrying out the principle of merit in societies that do not guarantee equal opportunities and that generate a feeling of scarce recognition and appreciation of those who receive lesser rewards: "In society's eyes, and perhaps also their own, their work no longer signified a valued contribution to the common good." [@sandel_tyranny_2020, pp.].

Empirical research on meritocracy has increased along with the philosophical-normative discussion on meritocracy in recent years. Particularly from a sociological perspective, meritocracy has been used in research on social mobility to characterize societies with low mobility that threaten the meritocratic ideal [@goldthorpe_myth_2003]. More recently, sociology and social psychology research has attended to the subjective aspects vis-a-vis beliefs in meritocracy. The label of beliefs in this realm covers a series of areas, such as attitudes, perceptions, and preferences (Castillo et al), whereby most of the link this subjective dimensions to individual socio-structural factors and  context-level determinants. For instance, some studies have analyzed how those with greater privileges believe more in meritocracy (Reynolds & Chan 2014), how greater economic inequality increases meritocratic beliefs [@mijs_paradox_2021a], and how larger inequality affects meritocratic beliefs (Morris et al 2022). Based on these findings, a research agenda has been reinforced on the legitimizing role of meritocracy, in line with previous studies using the concept of a just world (Lerner, Dalbert) and the theory of system justification (Jost & Major).

How do meritocratic beliefs legitimize inequalities? Empirical studies have used experiments and surveys to address this question. For instance, the evidence suggests that just world beliefs correlate negatively with support for redistributive compensation systems [@frank_performance_2015]. Conversely, individuals tend to support redistribution when they believe that the disadvantaged lack the opportunities to succeed [@evans_strong_2018]. @almas_cutthroat_2020 found that in a relatively unequal society (the United States), the highly educated accept inequality significantly more than the less educated because they perceive inequality as justifiable owing to differences in productivity (i.e., merit), whereas in a relatively equal society (Norway), the less educated accept inequality more, but not significantly more than the highly educated because meritocratic values are less prevalent. @barr_effect_2020 also addressed this triple interaction between the level of inequality in a society, the individual level of education, and the perceived origin of the disparity (either by luck or effort) to determine the extent to which inequality is accepted. They found that the interaction's mechanism varies depending on the compared societies. Finally, @garcia-sanchez_attitudes_2020, using data for 41 countries from the International Social Survey Programme (ISSP), found that the perceived size of the income gap correlated positively with support for progressive taxation. Still, this association was weaker among those who endorsed meritocratic and equal opportunity beliefs. In the same line, experimental research by Durante & Putterman (2009) subjects support less redistribution when the initial distribution is determined according to task performance.

Research about meritocratic beliefs at school age is rather scarce, leaving a wide research gap as schools are one of the primary socialization institutions where achievement based on merit explains success (Erivwo et al 2021). Nevertheless, it is possible to find some initial works focused on the area of distributive justice at school that are closely related to meritocratic beliefs, as the ones by Resh and Sabbagh (quotes). Using justice in grade obtention as a measure of distributive justice and meritocracy (Sabbagh et al 2006), they find for instance that a larger sense of distributive justice about grades is associated to higher socio-economic status [@resh_sense_2010], have a positive effect on liberal democratic orientation and on trust in people and in formal institutions (Resh & Sabbagh 2013), and tend to refrain from violence and to engage to a greater extent in extra-curricular school activity and community volunteering [@resh_sense_2017].
